---
output:
  pdf_document: default
bibliography: ["cytowania.bib"]
---

## Various data imputation techniques in R

*Authors: Jan Borowski, Filip Chrzuszcz, Piotr Fic  (Warsaw University of Technology)*

### Abstract

There are many suggestions how to deal with missing values in data sets problem. Some solutions are offered in publicly available packages for the R language. In our study, we tried to compare the quality of different methods of data imputation and their impact on the performance of machine learning models. We scored different algorithms on various data sets imputed by chosen packages. Results summary presents packages which enabled to achieve the best models predictions metrics. Moreover, duration of imputation was measured. 

### Introduction and Motivation

#### Background an related work

Missing observations in data sets is a common and difficult problem. In the field of machine learning, one of the key objects is the data set. Real-world data are often incomplete, which prevents the usage of many algorithms. Most implementations of machine learning models, available in popular packages, are not prepared to deal with missing values. Before creating a machine learning model, it is essential to solve the problem of missing observations. This requires user pre-processing of data. Some researches examined similarity between original and imputed data, in terms of descriptive statistics [@musil2002impcomp]. Missing data are common in medical sciences and impact of different imputation methods on analysis was measured [@bono2007medical]. Some studies show that imputation can improve the results of machine learning models and that more advanced techniques of imputation outperform basic solutions [@gustavo2003impclasif] [@su2008impclasif].    

#### Motivation

Various imputation techniques are implemented in different packages for the R language. Their performance is often analyzed independently and only in terms of imputation alone. Because of variety of available tools, it becomes uncertain which one package and method to use, when complete data set is needed for machine learning model. In our study we would like to examine, how methods offered by some popular packages perform on various data sets. We want to consider flexibility of these packages to deal with different data sets. The most important issue for us is the impact of performed imputation on later machine learning model performance. We are going to consider one specific type of machine learning tasks: *supervised binary classification*. Our aim is a comparison of metrics scores achieved by various models depending on chosen imputation method. 

#### Definition of missing data

At the beginning, clariyfing the definition of missing data is necessary. Missing data means, that one or more variables have no data values in observations. This can be caused by various reasons, which we can formally define as follows, referring to @rubin1976mar:

- MCAR (Missing completely at random)\
Values are missing completely at random if the events that lead to lack of value are independent both of observable variable and of unobservable parameters. The missing data are simply a random subset of the data. Analysis performed on MCAR data is unbiased. However, data are rarely MCAR.
- MAR (Missing at random)\
Missingness of the values can be fully explained by complete variables. In other words, missing data are not affected by their characteristic, but are related to some or all of observed data. This is the most common assumption about missing data.
- MNAR (Missing not at random)\
When data are missing not at random, the missingness is related to the characteristic of variable itself.

#### Techniques of dealing with missing data

In case of preparing a data set for machine learning models we can generally distinguish two approaches. The first method is **omission**. From the data set we can remove observations with at least one missing value or we can remove whole variables where missing values are present. This strategy is appropriate if the features are MCAR. However, it is frequently used also when this assumption is not met. It is also useless when the percentage of missing values is high. The second approach is **imputation**, where values are filled in the place of missing data. There are many methods of imputation, which we can divide into two groups. **Single imputation** techniques use information of one variable with missing values. Popular method is filling missings with mean, median or mode of no missing values. More advanced are predictions from regression models which are applied on the mean and covariance matrix estimated by analysis of complete cases. The main disadvantage of single imputation is treating the imputed value as true value. This method does not take into account the uncertainty of the missing value prediction. For this reason **multiple imputation** was proposed. This method imputes *k* values, which leads to creating *k* complete data sets. The analysis or model is applied on each complete data set and finally results are consolidated. This approach keeps the uncertainty about the range of values which the true value could have taken. Additionally, multiple imputation can be used in both cases of MCAR and MAR data.

### Methodology
Experiment like this one can be performed involving many techniques we decide to divide our tests into 4 steps:

- Data Preparation,
- Data Imputation,
- Model Training,
- Model Evaluation.

Bellow we will explain every step in detail.

#### 1. Data Preperation 
For test purposes we used 14 datasets form OpenML library [@R-OpenML]. Every dataset is designed for binary classification and most of them contain numerical and categorical features. Most of the sets have a similar number of observations in both classes but some of them were very unbalanced. Before data imputation data set was prepared specific preparation are different for each data set but we commonly do:

- Removing features which didn't contain useful information (for example all observation have the same value)
- Correcting typos and converting all string to lower case to reduce the number of categories 
- Converting date to more than one column (for example "2018-03-31" can be converted to three column year, month and day)
- Removing or converting columns with too many categories

After cleaning datasets were transferred to the next step. 

#### 2.Data Imputation
Clean datasets were split into two data sets train and test in proportion $1/4$ respectively. This split was performed randomly and only once for every dataset that's mean every imputation technique used the same split. Imputation was performed separately for train and test sets. Before split we also remove the target column to avoid using it in imputation. For our study, we decided to choose five packages designed for missing data imputation in the R language and one self-implemented basic technique:

- **Mode and median**: Simple technique of filling missing values with mode (for categorical variables) and median (for continuous variables) of complete values in a variable. Implemented with basic R language functions.
- **mice**[@mice2011]: Package allows to perform multivariate imputation by chained equations (MICE), which is a type of multiple imputation. The method is based on Fully Conditional Specification, where each incomplete variable is imputed by separate model.
- **missMDA**[@missMDA2016]: Package for multiple missing values imputation. Data sets are imputed with the principal component method, regularized iterative FAMD algorithm (factorial analysis for mixed data). First estimation of the number of dimensions for factorial analysis is essential.
- **missFOREST**[@missForest2012]: Package can be used for imputation with predictions of random forest model, trained on complete observations. Package works on data with complex interactions and non-linear relations. Enables parallel calculations.
- **softImpute**[@R-softImpute]: Package for matrix imputation with nuclear-norm regularization.  Algorithm works like EM, solving an optimization problem using a soft-thresholded SVD algorithm.  Works only with continuous variables.
- **VIM**[@VIM2016]: Package for visualization and imputation of missing values. It offers iterative robust model-based imputation (IRMI). In each iteration, one variable is used as a response variable and the remaining variables as the regressors.


First we use mode/median imputation, which is a very simple method and it is used as a base result for more complex algorithms to compare.
Imputation method form mice package don't require any form of help because can impute numeric and categorical features.
SoftImpute package works only with numeric features. To compare it with other algorithms on the same data we use SoftImput for numeric variables and mode for categorical variables. Alternatively, it is possible to use SoftImpute for numeric features and different algorithms for categorical variables, but we decided that this approach may lead to unreliable results.
MissForest algorithm can be used on both numeric and categorical features and is capable of performing imputation without any help of other methods.
Imputation method from Mice package also can be run on all types of data.
Iterative Robust Model-Based Imputation method from VIM package can impute all types of data. This method additionally creates new columns with information whether observation was imputed or not. We decided to do not use these columns, because other methods do not create them.
Last method which we covered is missMDA which also can be used to input numeric and categorical features.
After imputation we add back target variable to both sets. All methods work on the same parameters for all datasets. In case when for some reason method can't input some dataset it was treated like "worst result" more information about it can be found in section 4.Model evaluation.

#### 3.Model traing 
For classification task we use four classification algorithms:

- Extreme Gradient Boosting, 
- Random Forest, 
- Support Vector Machines,
- Linear Regression 

All methods were implemented in **mlr** package [@mlr] for hyperparameter tuning we also use methods form the dame package. For all data sets four classifications were trained and tuned on the same train sets. To select params we used Grid Search. We will not focus on this part of the experiment. The most important part in this step is that every imputation was proceeded the same whats mean different in results can be caused  
only by different imputation Technik. 

#### 4.Model evaluation 
After previous steps we have trained models and test sets. In the final step we evaluate model and imputation. For every imputation and algorithm we calculate F1 score expres by formula $2\frac{(precision)*(recall)}{precision + recall}$ and accuracy. In case when imputation algorithm fail to impute some dataset results for this set are thread as "the worst result" it means if u try to carte a ranking it is always last (if more then one imputation fail all of them is placed last). A detailed discussion about results in the next section.


 
### Results

After the long and tedious process of data imputation it is finally time to evaluate our methods of imputation. Methods were trained and tested on 11 datasets and evaluated strictly using these 4 methods mentioned earlier. Before score analysis it is worth mentioning that all algorithms were tested with basic parametres, so there is certainly some room for improvementm but for now we decided to leave these scores as they are. Table below presents average metrics achieved by model, depending on imputation method.
```{r echo=FALSE, message=FALSE, warning=FALSE , results='asis'}
library(knitr)
library(tidyverse)
scores <- readr::read_csv('./wyniki_csv/wyniki_moje.csv')

ramka <- scores %>% group_by(Method) %>% summarise(accuracy = round(mean(Acc),2),
                                          auc = round(mean(Auc),2),
                                          precision = round(mean(Precision),2),
                                          recall = round(mean(Recall),2))

knitr::kable(ramka,caption='Average scores of model on imputation methods')
```

Our experiment definately did not find out the best imputation algoriithm, but we can derive some interesting conclusions from it.
First of all, we are going to treat median imputation as kind of a baseline for our algorithms. Suprisngly it seems to perform quite well, amongst the others, often quite sophisticated methods. It achieved over 80% of accuracy on average. However, it is hard to decide whether it is a high score or not, because we are only able to compare algorithms between the others. To say anything more meanigful about our methods we shall look at the distribution of the scores in order to make a better analysis of performance.\


#### ROC AUC score

As mentioned earlier we decided to use a few metrics to compare algorithms. One of it is ROC AUC score. It is a measure which caluclates TPR (which is ratio beetwen our predicted positive examples over overall amount of positive examples) and FPR (which is ratio beetwen false positives and all negative examples). Next TPR and FPR are evaluated at different thershold values and than we have something called ROC curve. Our measure is simply the area under that curve. Obviously the maximum of that score is 1. ROC AUC is also quite good for working with imbalanced sets, which may be useful for us, because we had to deal with such in our experiment.
```{r echo=FALSE, message=FALSE, warning=FALSE}
p<-ggplot(scores, aes(x=Method, y=Auc, color=Method)) +
  geom_boxplot()+theme_minimal()+labs(x="",y="")+ggtitle("ROC AUC among different imputations")
p
```

Taking a brief look at the distributions of the scores it is clear that median is not so good. Its scores seems to vary the most among datasets, which is not something we ideally want. Also it is worth noting that unfortunately missMDA achieved the worse score and also was the most problematic imputation technic from which we tested. It did not manage the perform its imputations on 5 out of 11 datasets, which clearly disqualify this method from everyday usage. Also softimpute had problem with working on all sets, albeit it managed to perform quite decent scores. On the other end are VIM irmi, missFOREST and Mice. All of these three packages managed to make imputations on all sets, and achieved good scores.

#### Precision and recall
Moving further into score analysis we need to focus on two undeniably important parameters, which are precision and recall. We can understand recall as an ability of a model to find all the relevant cases within a dataset, while precision  expresses the proportion of the data our model says was relevant actually were relevant. Ideally we would like to make both of these metrics as high, as possible, which is of course not so possible. Instead we can focus on one of these two depending what problem we are solving. However, in our experiment we are not capable to say which metric is more important for us, so we can simply compare them, and on that basis say something about our methods.

```{r echo=FALSE, message=FALSE, warning=FALSE,fig.width=4.5,fig.height=5,fig.show='hold'}
p<-ggplot(scores, aes(x=Method, y=Precision, color=Method)) +
  geom_boxplot()+theme_minimal()+labs(x="",y="")+ggtitle("Precision among different imputations")
p1<-ggplot(scores, aes(x=Method, y=Recall, color=Method)) +
  geom_boxplot()+theme_minimal()+labs(x="",y="")+ggtitle("Recall among different imputations")
p
p1
```

We can now focus on brief analysis of the best scores, which our methods managed to achieve. In terms of precision VIM irmi, missFOREST and Mice achieved almost the same scores, with Mice having the most compact scores. Despite that scores quite comparable with these which we have seen on ROC AUC scores, so it will be more interesting to move on for recall scores. Here once again VIM irmi leads with the highest median of scores, however ironically median imputing has the second median of scores here. After looking at its scores, we can derive our first bold conclusion. Namely it is worth to use median imputation when we aim for decent scores and do not care about dispersion of it. Quite simmilary to the previous scores mice and missFOREST were close to the leaders and softImpute and missMDA were at the back of the field.

#### Time of imputations

The last topic we wish to consider in our analysis section is the time of imputations. It is quite significant, because even the best imputation may become redundant if it has too long working time.

```{r echo=FALSE, message=FALSE, warning=FALSE,fig.width=4.5,fig.height=5,fig.show='hold'}
p<-ggplot(scores, aes(x=Method, y=Time, color=Method)) +
  geom_boxplot()+theme_minimal()+labs(x="",y="")+ggtitle("Time in seconds ")
p
```

We can divide our algorithms into two groups. In the first one, there are:\
- VIM irmi\
- Median\
- Mice\
- SoftImpute\
This group is simply characterised by very low time need to peroform the imputation. Second group consists of: \
- missFOREST\
- missMDA\
Algorithms from that group had that quite high, which may be unacceptable while working with huge datasets.

### Summary and conclusions 

Considering metrics scores achieved by evaluated model few important conlusions about imputation methods can be made. Basic approach with median and mode imputation allows to get decent results of machine learning model. Its disadvatage is large discrepancy in scores. More complex techniques available in packages, ensure more stable results of model among all data sets. From them, we can distinguish missFOREST, VIM irmi and mice. These packages have carried out the best imputation in terms of further model performance. They proved to be the most versatile, achiving the best average results of accuracy. VIM and missFOREST AUC scores show also the stability of model performance on imputed data set. SoftImpute package caused too low scores of recall and precision which suggests its poor performance on unbalanced data. MissMDA package performed worse than basic mode/median and all other methods, being able to proceed only on 6 from 11 data sets.  
Second aspect to consider is time needed for imputation. MissFOREST achieved good metrics scores of model, but its very time consuming. Both VIM irmi and mice packages not only provided well imputed data sets but also were time efficient.  
Choosing the best imputation tool may obviously depend on characteristic of the specific data sets, however our research should help to find flexible solution for everyday usage.

### Bibliography